---
title: "step1.1 generate data (first three criteria)"
output: pdf_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, warning= F, message = F)
rm(list=ls(all=TRUE))
library(readstata13)
mainDir <- "/Users/oliviazhang/Desktop/"
```


## NHATS 2011 cohort
```{r}
#Read in NHATS full data
Round_1_FULL <- read.dta13(paste0(mainDir,"/MRP/NHATS_R1_Final_Release_STATA_V3/NHATS_Round_1_SP_File.dta"))
```


## Select relevant variables
```{r}
selected_svy <- c("spid",
              "w1anfinwgt0", 
              "w1varunit", #cluster
              "w1varstrat") #strata

selected_frail <- c("hw1currweigh","hw1trytolose","hw1lst10pnds","hw1howtallft","hw1howtallin",
                    "ss1lowenergy","ss1loenlmtat",
                    "wa1wlkc1secs","wa1wlk1hndr", "wa1rsn11wkc", "wa1rsn21wkc", "wa1rsn31wkc", "wa1wlkc1rslt", "wa1dwlkadm",
                    "gr1grp1rdng","gr1grp2rdng", "gr1grp1noat1", "gr1grp1noat2", "gr1grp1noat3", "gr1grp1reslt", "gr1dgripadm",
                    "pa1evrgowalk","pa1vigoractv")
selected_demo <- c("r1dgender","rl1dracehisp","r1d2intvrage","r1dresid","re1dcensdiv")

Round_1_data <- Round_1_FULL[,c(selected_frail,selected_svy,selected_demo)]
length(Round_1_data$spid)
```

## Only community-living participants
```{r}
Round_1_data <- Round_1_data[(Round_1_data$r1dresid == 1 | Round_1_data$r1dresid == 2), ]
length(Round_1_data$spid)
```

```{r}
Round_1_data$gender <- as.factor(Round_1_data$r1dgender)
```



## Frailty criteria #1: weight loss

Participants met this criteria if:
1. they had body mass index (BMI) less than 18.5 kg/m2, 
2. or reported unintentionally losing 10 or more pounds in the last year.

### Step 1: Calculate BMI:

```{r}
#YOU CURRENTLY WEIGH
table(Round_1_data$hw1currweigh[Round_1_data$hw1currweigh<0], useNA="ifany") #-8 DK, -7 RF
#HOW TALL ARE YOU FEET
table(Round_1_data$hw1howtallft[Round_1_data$hw1howtallft<0], useNA="ifany") #-8 DK, -7 RF
#HOW TALL ARE YOU INCHES 
table(Round_1_data$hw1howtallin[Round_1_data$hw1howtallin<0], useNA="ifany") #-8 DK, -1 Inapplicable
```


NHATS variables used for BMI:
```{r}
#lbs to kg
Round_1_data$hw1currweigh <- ifelse(Round_1_data$hw1currweigh < 0 , NA, Round_1_data$hw1currweigh) 
Round_1_data$weight <- Round_1_data$hw1currweigh*0.453592

#ft to inch
Round_1_data$hw1howtallft <- ifelse(Round_1_data$hw1howtallft < 0 , NA, Round_1_data$hw1howtallft) 
Round_1_data$hw1howtallin <- ifelse(Round_1_data$hw1howtallin < 0 , NA, Round_1_data$hw1howtallin) 

#inch to cm
Round_1_data$height <- Round_1_data$hw1howtallft*.3048 + Round_1_data$hw1howtallin*.0254
```


```{r}
Round_1_data$bmi <- Round_1_data$weight/((Round_1_data$height)^2)
Round_1_data$bmi[is.na(Round_1_data$bmi)] <- -1
summary(Round_1_data$bmi)
```



### Step 2: Generate new variable for wtloss:

NHATS variables used:
```{r}
#WERE YOU TRYNG LOSE WEGHT
table(Round_1_data$hw1trytolose, useNA="ifany") #-1 Inapplicable, 1 YES, 2 NO
#LOST 10 POUNDS IN LAST YR
table(Round_1_data$hw1lst10pnds, useNA="ifany") #-8 DK, -7 RF, 1 YES, 2 NO
```


```{r}
Round_1_data$wtloss <- ifelse((Round_1_data$bmi >= 0 & Round_1_data$bmi < 18.5) | (Round_1_data$hw1lst10pnds == 1 & Round_1_data$hw1trytolose == 2), 1,
                               ifelse(Round_1_data$bmi >= 18.5 & (Round_1_data$hw1lst10pnds == 2 | (Round_1_data$hw1lst10pnds == 1 & Round_1_data$hw1trytolose == 1)), 0, NA))
                                           
table(Round_1_data$wtloss, useNA="ifany")
table(Round_1_data$wtloss, useNA="ifany")[2]/nrow(Round_1_data)
```


## Frailty criteria #2: exhaustion

Participants met this criteria if:
1. reported recently having low energy
2. AND enough to limit their activities


NHATS variables used:
```{r}
#LOW ENERGY IN LAST MONTH
table(Round_1_data$ss1lowenergy, useNA="ifany") #ss1lowenergy: -8 DK, 1 Yes, 2 No
#LOW ENERGY EVER LIM ACT
table(Round_1_data$ss1loenlmtat, useNA="ifany") #ss1loenlmtat: -8 DK, -1 Inapplicable, 1 Yes, 2 No
```


```{r}
Round_1_data$exhaust <- ifelse(Round_1_data$ss1lowenergy == 1 & Round_1_data$ss1loenlmtat == 1, 1,
                               ifelse(Round_1_data$ss1lowenergy == 2 | (Round_1_data$ss1lowenergy == 1 & Round_1_data$ss1loenlmtat == 2), 0, NA))

table(Round_1_data$exhaust, useNA="ifany")
table(Round_1_data$exhaust, useNA="ifany")[2]/nrow(Round_1_data)
```

\newpage 

## Frailty criteria #3: low energy expenditure

Participants met this criteria if: 

1. recently, they never walked for exercise 
2. or never engaged in vigorous activities

NHATS variables used:
```{r}
#EVER GO WALKING
table(Round_1_data$pa1evrgowalk, useNA="ifany") #-8 DK, -7 RF, 1 YES, 2 NO
#EVER VIGOROUS ACTIVITIES
table(Round_1_data$pa1vigoractv, useNA="ifany") #-8 DK, -7 RF, 1 YES, 2 NO
```

Generate new variable for lowenergy:
```{r}
Round_1_data$lowenergy <- ifelse(Round_1_data$pa1evrgowalk == 2 & Round_1_data$pa1vigoractv == 2, 2,
                                 ifelse(Round_1_data$pa1evrgowalk == 1 & Round_1_data$pa1vigoractv == 1, 0,
                                        ifelse((Round_1_data$pa1evrgowalk == 1 & Round_1_data$pa1vigoractv == 2) | (Round_1_data$pa1evrgowalk == 2 & Round_1_data$pa1vigoractv == 1), 1, NA)))

table(Round_1_data$lowenergy, useNA="ifany")
table(Round_1_data$lowenergy, useNA="ifany")[3]/nrow(Round_1_data)
```

\newpage 

## Frailty criteria #4: low walking speed

“Low walking speed” was defined, using the first of two usual-pace walking trials, as being at or below the 20th percentile of the weighted population distribution within four sex-by-height categories. 


NAHTS variables used:
```{r}
table(Round_1_data$wa1wlkc1secs[Round_1_data$wa1wlkc1secs<0], useNA="ifany") #-9 Missing, -1 Inapplicable
table(Round_1_data$wa1wlk1hndr[Round_1_data$wa1wlk1hndr<0], useNA="ifany") #-9 Missing, -1 Inapplicable, -8 DK, -7 RF
```


```{r}
Round_1_data$wa1wlkc1secs <- ifelse(Round_1_data$wa1wlkc1secs < 0 , NA, Round_1_data$wa1wlkc1secs)
Round_1_data$wa1wlk1hndr <- ifelse(Round_1_data$wa1wlk1hndr < 0 , NA, Round_1_data$wa1wlk1hndr)

Round_1_data$sec <- Round_1_data$wa1wlkc1secs + Round_1_data$wa1wlk1hndr/100
Round_1_data$speed <- 3/Round_1_data$sec
summary(Round_1_data$speed)
```


```{r}
#replace missing with 0 for participants who didn't complete the test for safety reasons
for (i in 1:nrow(Round_1_data)){
  if (is.na(Round_1_data$sec[i])){
      if (Round_1_data$wa1wlkc1rslt[i] == 2 | Round_1_data$wa1dwlkadm[i] == 4 | Round_1_data$wa1rsn11wkc[i] == 1 | Round_1_data$wa1rsn21wkc[i] == 1 | Round_1_data$wa1rsn31wkc[i] == 1){
    #-9 Missing, -1 Inapplicable, 2 Attempted, 3 Not attempted,
    Round_1_data$speed[i] <- 0
  }
}
}
#Remove outlier (same as Bandeen-Roche's code)
Round_1_data$speed[Round_1_data$speed >10 & !is.na(Round_1_data$speed)] <- NA

summary(Round_1_data$speed)
summary(Round_1_data$speed)[7]/nrow(Round_1_data)
```


\newpage

## Frailty criteria #5: low grip strength

“Weakness” was defined, using maximum dominant hand grip strength over two trials, as being at or below the 20th percentile within eight sex-by-BMI categories. 


NAHTS variables used:
```{r}
table(Round_1_data$gr1grp1rdng[Round_1_data$gr1grp1rdng<0], useNA="ifany") #-9 Missing, -1 Inapplicable
table(Round_1_data$gr1grp2rdng[Round_1_data$gr1grp2rdng<0], useNA="ifany") #-9 Missing, -1 Inapplicable
```


```{r}
Round_1_data$gr1grp1rdng <- ifelse(Round_1_data$gr1grp1rdng < 0 , NA, Round_1_data$gr1grp1rdng)
Round_1_data$gr1grp2rdng <- ifelse(Round_1_data$gr1grp2rdng < 0 , NA, Round_1_data$gr1grp2rdng)
```


```{r}
#Find maximum dominant hand grip strength over two trials
Round_1_data$grip <- NA
Round_1_data$gripi <- NA

for (i in 1:nrow(Round_1_data)){
  #if not NA for both trial
  if (!is.na(Round_1_data$gr1grp1rdng[i]) & !is.na(Round_1_data$gr1grp2rdng[i])){
      if (Round_1_data$gr1grp1rdng[i] < Round_1_data$gr1grp2rdng[i]){
        Round_1_data$grip[i] <- Round_1_data$gr1grp2rdng[i]
        Round_1_data$gripi[i] <- 2
        }
    else {
      Round_1_data$grip[i] <- Round_1_data$gr1grp1rdng[i]
      Round_1_data$gripi[i] <- 1
      }
  }
  #if NA in one trial
  if (is.na(Round_1_data$gr1grp1rdng[i])){
    Round_1_data$grip[i] <- Round_1_data$gr1grp2rdng[i]
    Round_1_data$gripi[i] <- 2
  }
  if (is.na(Round_1_data$gr1grp2rdng[i])){
    Round_1_data$grip[i] <- Round_1_data$gr1grp1rdng[i]
    Round_1_data$gripi[i] <- 1
  }
}

summary(Round_1_data$grip)
summary(Round_1_data$grip)[7]/nrow(Round_1_data)
table(Round_1_data$gripi, useNA="ifany")
```


```{r}
#replace missing with 0 for participants who didn't complete the test for safety reasons
for (i in 1:nrow(Round_1_data)){
  if (is.na(Round_1_data$grip[i])){
    if (Round_1_data$gr1grp1reslt[i] == 2 | Round_1_data$gr1dgripadm[i] == 4 | Round_1_data$gr1grp1noat1[i] == 1 | Round_1_data$gr1grp1noat2[i] == 1 | Round_1_data$gr1grp1noat3[i] == 1){
    #-9 Missing, -1 Inapplicable, 2 Attempted, 3 Not attempted,
    Round_1_data$grip[i] <- 0
    }
}
}

#grip strength > 60 women and >90 for men are outliers
Round_1_data$grip[Round_1_data$grip >90 & !is.na(Round_1_data$grip) & Round_1_data$r1dgender == 1] <- NA
Round_1_data$grip[Round_1_data$grip >60 & !is.na(Round_1_data$grip) & Round_1_data$r1dgender == 2] <- NA

summary(Round_1_data$grip)
summary(Round_1_data$grip)[7]/nrow(Round_1_data)
```







### Generate new variable for sex-by-height categories:
```{r}
Round_1_data$height[is.na(Round_1_data$height)] <- -1
table(Round_1_data$height)
```


```{r}
Round_1_data$ht_gender <- "Height_missing"

Round_1_data$ht_gender[Round_1_data$gender == 2 & Round_1_data$height <= 1.59 & Round_1_data$height >=0] <- "Female, 1"
Round_1_data$ht_gender[Round_1_data$gender == 2 & Round_1_data$height > 1.59] <- "Female, 2"

Round_1_data$ht_gender[Round_1_data$gender == 1 & Round_1_data$height <= 1.73 & Round_1_data$height >=0] <- "Male, 1"
Round_1_data$ht_gender[Round_1_data$gender == 1 & Round_1_data$height > 1.73] <- "Male, 2"

table(Round_1_data$ht_gender, useNA = "ifany")
```



### Generate new variable for sex-by-BMI categories:
```{r}
Round_1_data$bmi_gender <- "BMI_missing"
Round_1_data$bmi_gender[Round_1_data$gender == 2 & Round_1_data$bmi >= 0 & Round_1_data$bmi < 18.5] <- "Female, 1"
Round_1_data$bmi_gender[Round_1_data$gender == 2 & Round_1_data$bmi >= 18.5 & Round_1_data$bmi < 25] <- "Female, 2"
Round_1_data$bmi_gender[Round_1_data$gender == 2 & Round_1_data$bmi >= 25 & Round_1_data$bmi < 30] <- "Female, 3"
Round_1_data$bmi_gender[Round_1_data$gender == 2 & Round_1_data$bmi >= 30] <- "Female, 4"

Round_1_data$bmi_gender[Round_1_data$gender == 1 & Round_1_data$bmi >= 0 & Round_1_data$bmi < 18.5] <- "Male, 1"
Round_1_data$bmi_gender[Round_1_data$gender == 1 & Round_1_data$bmi >= 18.5 & Round_1_data$bmi < 25] <- "Male, 2"
Round_1_data$bmi_gender[Round_1_data$gender == 1 & Round_1_data$bmi >= 25 & Round_1_data$bmi < 30] <- "Male, 3"
Round_1_data$bmi_gender[Round_1_data$gender == 1 & Round_1_data$bmi >= 30] <- "Male, 4"

table(Round_1_data$bmi_gender, useNA="ifany")
```



### Count the number of missing scales
Persons assessed on at least three frailty criteria (including “0” scores) were considered eligible
```{r}
Round_1_data$misscnt <- -1
for (i in 1:nrow(Round_1_data)){
  a <- as.numeric(!is.na(Round_1_data$wtloss[i]))
  b <- as.numeric(!is.na(Round_1_data$exhaust[i]))
  c <- as.numeric(!is.na(Round_1_data$lowenergy[i]))
  d <- as.numeric((!is.na(Round_1_data$speed[i]) & !(Round_1_data$ht_gender[i]=="Height_missing")))
  e <- as.numeric((!is.na(Round_1_data$grip[i]) & !(Round_1_data$bmi_gender[i]=="BMI_missing")))
  validcnt <- a + b + c + d + e
  Round_1_data$misscnt[i] <- 5-validcnt
}
table(Round_1_data$misscnt,useNA="ifany")
round(table(Round_1_data$misscnt,useNA="ifany")/nrow(Round_1_data)*100,1)
```



```{r}
save.dta13(Round_1_data,paste0(mainDir,"/MRP/2011_cohort/data/cleaned_data_7609.dta"))
```



